## Setting Up

We'll use the `midd_survey` data for this example. Load it and the usual packages.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(pander)

midd_survey <- read.csv("https://raw.githubusercontent.com/mjclawrence/soci385/master/data/midd_survey.csv")
```

## Formatting Regression Tables With Stargazer

Summaries of linear regression models have a lot of information and can be messy in knitted files. One option to clean them up is to use the `stargazer` package. The package is highly customizable; we'll go over just a few options today. For more details about the package, click [here](https://www.jakeruss.com/cheatsheets/stargazer/). 

To see how stargazer compares to the output we are used to, use the `midd_survey` data to regress `gpa` on `gender`, save the model as `model1`, and look at the summary (remember `echo = FALSE` suppresses the code in the knitted file):

```{r, echo = FALSE}
model1 <- lm(gpa ~ siblings, data = midd_survey)

summary(model1)
```

Now that you are comfortable interpreting summary outputs, you should continue using the traditional summary outputs to get a sense of your models. But replace `summary()` with `stargazer()` when you want to knit your file. 

```{r}
# INSTALL STARGAZER THE FIRST TIME FROM THE BOTTOM RIGHT PANE
library(stargazer)

stargazer(model1)
```

There are a few other things to add to make sure the file knits properly:

- `results = 'asis'` is the most important thing. It leaves the output in the format that knitr needs to render it. (Add this option to the open fence line; add everything else to the stargazer function line.)
- `header = FALSE` deletes the contact information for the package author
- `type = 'latex'` converts the output into the Latex language that pandoc needs to knit to a PDF. Latex is the default so you can leave it off if you want. To change to web output you would want to change `type = 'latex'` to `type = 'html'`.

```{r, results = 'asis', warning = FALSE, echo = FALSE}
stargazer(model1, header = FALSE, type = 'latex')
```

Knit the file to see the formatted regression table. 

\pagebreak

This is better than dumping raw R output into a report, but we can do even better. 

Here are some more changes for the `stargazer()` function:

- `style = 'ajs'` formats the table following publication guidelines for the *American Journal of Sociology*. Replace 'ajs' with 'asr' to use the formatting style for the *American Sociological Review*. These are the two most common formatting styles you will see for sociology papers. I prefer ajs style for written reports (because it includes standard errors), and asr style for presentation slides (because it is more condensed).
- `title` creates a title for the table which will be numbered automatically
- `covariate.labels = c()` is a list of names for the independent and control variables in your model. You need to label every variable or this won't work.
- `dep.var.labels = ` creates a title above the dependent variable model numbers
- `keep.stat = c("n", "rsq")` reports the number of observations and the r-squared value but excludes all the other model information

We can include more than one model in a table too. Let's save a model with a control variable and a model with an interaction.

```{r, include = FALSE}
model2 <- lm(gpa ~ siblings + gender, data = midd_survey)
model3 <- lm(gpa ~ siblings * gender, data = midd_survey)
```

And then include our three models in a stargazer function that also includes several of the changes described above.

```{r, results = 'asis', warning = FALSE}
stargazer(model1, model2, model3,
          header=FALSE, type = 'latex', style = 'asr',
  title = "A Better Title",
  covariate.labels = c("Number of Siblings", "Gender = Other", "Gender = Woman",
                       "Siblings X Gender = Other", "Siblings X Gender = Woman"),
  dep.var.labels = "GPA",
  keep.stat = c("n", "rsq"))
```


